---
title: "STAT 526 HW 8"
author: "Bowen Zheng"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem 0

Bowen Zheng

# Problem 1

> The log-log confidence interval for S(t) is often recommended. Given that $Var(log[S(t)]) \approx \sum \frac{d_j}{(n_j - d_j) n_j}$ use the delta method to approximate the variance of log(-log[S(t)]). Make sure to show all the calculation steps.

We start with our random variable being $Y = \log[\hat{S}(t)]$ and then are transforming it with log, ie finding $\log(-\log[\hat{S}(t)])$. For the delta method, we know $\text{Var}(g(Y)) \approx [g'(\mu_Y)]^2 \text{Var}(Y)$.

Then $g(Y) = \log(-Y)$.

We calculate the first derivative:
$$g'(Y) = \frac{d}{dY} \log(-Y) = \frac{1}{-Y} \cdot (-1) = \frac{1}{Y}$$
Which means:
$$g'(\log[\hat{S}(t)]) = \frac{1}{\log[\hat{S}(t)]}$$

Now we plug into our delta method equation:
$$\widehat{\text{Var}}(\log(-\log[\hat{S}(t)])) \approx \left[ \frac{1}{\log[\hat{S}(t)]} \right]^2 \widehat{\text{Var}}(\log[\hat{S}(t)]) = \frac{1}{[\log\hat{S}(t)]^2} \sum_{t_j \le t} \frac{d_j}{(n_j - d_j)n_j}$$


\newpage

# Problem 2

> The survival times of 11 patients with Acute Myelogenous Leukemia are given below: 9, 13, 13+, 18, 23, 28+, 31, 34, 45+, 48, 161+, where censored data are marked by “+” in the superscript.

## a.)

> Calculate the Kaplan-Meier estimate of S(20).

Our KM estimate:
$$\hat{S}(t) = \prod_{i: t_i \le t} \left( 1 - \frac{d_i}{n_i} \right)$$


| $t_i$ | $n_i$ (At Risk) | $d_i$ (Events) | $1 - d_i/n_i$ | $\hat{S}(t_i)$ |
| :--- | :--- | :--- | :--- | :--- |
| 9 | 11 | 1 | 10/11 | 0.9091 |
| 13 | 10 | 1 | 9/10 | 0.8182 |
| 18 | 8 | 1 | 7/8 | 0.7159 |

Next death at t=23 so we do not need to calculate that far. Our final value is 0.7159 as our estimate for S(20).

## b.)

> Obtain the standard error of $log \hat{S}(20)$.

We use the identity given to us earlier. 
$$Var(\log \hat{S}(t)) = \sum_{i: t_i \le t} \frac{d_i}{n_i(n_i - d_i)}$$

Looking at our table earlier, we can grab these numbers for the time events 9,13,18. We have $t=9$: $\frac{1}{11(10)} = 0.00909$, $t=13$: $\frac{1}{10(9)} = 0.01111$, $t=18$: $\frac{1}{8(7)} = 0.01786$.

Then $Var(\log \hat{S}(20)) = 0.00909 + 0.01111 + 0.01786 = 0.03806$ and $SE(\log \hat{S}(20)) = \sqrt{0.03806} \approx \mathbf{0.1951}$.

## c.)

> Assuming an exponential distribution, estimate S(20).

We know $S(t) = e^{-\lambda t}$. MLE for $\lambda$ is $\hat{\lambda} = \frac{D(t)}{\sum T_i}$. 

We can calculate $\hat{\lambda} = \frac{7}{423} \approx 0.01655$ and then $\hat{S}(20) = e^{-0.01655 \times 20} = e^{-0.331} \approx 0.7182$

## d.)

> Obtain the standard error of $log \hat{S}(20)$ for the exponential estimate

We know $\log S(t) = -\lambda t$ under exponential modeling. 

If we assume it is exponential then variance of $\hat{\lambda}$ is $Var(\hat{\lambda}) = \frac{\lambda^2}{D}$.

Combine these and we can calculate the variance:
$$Var(\log \hat{S}(t)) = Var(-\hat{\lambda} t) = t^2 Var(\hat{\lambda}) = \frac{t^2 \hat{\lambda}^2}{D}$$

Now we plug the values we got in part c in:
$$Var(\log \hat{S}(20)) = \frac{20^2 \times (0.01655)^2}{7} = \frac{400 \times 0.0002739}{7} \approx 0.01565$$

Take the square root and we get $SE(\log \hat{S}(20)) = \sqrt{0.01565} \approx 0.1251$



\newpage


# Problem 3

> The following data report the survival of 30 late stage liver cancer patients (in weeks) according to three different treatment methods. In Homework #7, you tested for differences among the treatments nonparametrically using the log-rank test and parametrically assuming the survival times were exponential.

## a.)

> Test whether the exponential or Weibull distributions best fit the data.

Use libraries Prof Song uses.


```{r}
library(survival)
library(survminer)

# Create the dataset
time <- c(2, 2, 4, 6, 7, 9, 12, 17, 19, 24, # Trt 1
          1, 1, 3, 4, 6, 6, 8, 9, 10, 12, # Trt 2
          5, 6, 8, 10, 12, 13, 13, 14, 19, 21) # Trt 3

cens <- c(0, 1, 1, 0, 1, 1, 1, 0, 0, 1, # Trt 1
            1, 1, 1, 0, 1, 1, 0, 1, 1, 0, # Trt 2
            0, 1, 1, 1, 0, 1, 1, 0, 1, 1) # Trt 3

trt <- factor(rep(1:3, each = 10))
surv_obj <- Surv(time, cens)

fit_weibull <- survreg(surv_obj ~ trt, dist = "weibull")
fit_exponential <- survreg(surv_obj ~ trt, dist = "exponential")
summary(fit_weibull)
summary(fit_exponential)
```

Use simple likelihood ratio test to test if exponentional model (special version of weibull) is better.

```{r}
L_weibull <- logLik(fit_weibull)
L_exp <- logLik(fit_exponential)
LR_stat <- as.numeric(2 * (L_weibull - L_exp))
pchisq(LR_stat, df = 1, lower.tail = FALSE)
```

Since the p-value is less than $\alpha = 0.05$, we reject the exponential model in favor of the weibull model. 

## b.)

> Assuming the survival times are Weibull, what are the mean survival lives for each group?

We know that the expected value of Weibull distribution is $E[T] = \lambda \Gamma\left(1 + \frac{1}{k}\right)$ with $k = 1 / \text{scale}$ (from the survreg output) and $\lambda = \exp(\text{intercept} + \beta_i)$.

We get these values from our models above. 

```{r}
coeffs <- coef(fit_weibull)
shape_k <- 1 / fit_weibull$scale

# Calculate means for each treatment group
# Trt 1 (intercept/baseline)
mean_trt1 <- exp(coeffs[1]) * gamma(1 + 1/shape_k)

# Trt 2
mean_trt2 <- exp(coeffs[1] + coeffs[2]) * gamma(1 + 1/shape_k)

# Trt 3
mean_trt3 <- exp(coeffs[1] + coeffs[3]) * gamma(1 + 1/shape_k)

mean_trt1
mean_trt2
mean_trt3
```

Our mean survival times are 14.31066 for treatment 1, 7.413575 for treatment 2, 14.09647 for treatment 3. 

## c.)

> The Cox proportional hazard model was fit giving the following output. The estimates of survival functions for the first treatment group are $\hat{S}(1) = 0.9612$ and  $\hat{S}(2) = 0.9405$. Compute the estimates of the survival functions of the corresponding values for the second and third groups, respectively.

Treatment group 1 is the baseline (intercept) and we assume by the proportional hazard model that the thing that changes is by slope not affected by time, ie:
$$S(t | x) = [S_0(t)]^{\exp(\beta x)}$$

$\hat{S}_1(t)$ is our estimate for the baseline survival $S_0(t)$,  $\beta_2 = 3.13901$ and $\beta_3 = 1.07694$. 

We can now calculate:

At $t = 1$:

  - Group 2: $\hat{S}_2(1) = [\hat{S}_1(1)]^{HR_2} = (0.9612)^{3.13901} \approx \mathbf{0.8837}$
  - Group 3: $\hat{S}_3(1) = [\hat{S}_1(1)]^{HR_3} = (0.9612)^{1.07694} \approx \mathbf{0.9583}$

At $t = 2$:

  - Group 2: $\hat{S}_2(2) = [\hat{S}_1(2)]^{HR_2} = (0.9405)^{3.13901} \approx \mathbf{0.8242}$
  - Group 3: $\hat{S}_3(2) = [\hat{S}_1(2)]^{HR_3} = (0.9405)^{1.07694} \approx \mathbf{0.9360}$

## d.)

> Provide a graphical assessment to diagnose whether a Cox proportional hazard model can be used

Now we plot the log-log survival plot. We plot log of the $\Lambda(t)$ vs log of the time. This is because the assumption is $log (\Lambda(t)) = log(-log S(t))$ is linear. 

```{r}
# 1. Fit the Kaplan-Meier curves for each treatment group
km_fit <- survfit(Surv(time, cens) ~ trt)

# 2. Create the log-log plot
# note: fun = "cloglog" transforms S(t) to log(-log(S(t)))
plot(km_fit, 
     fun = "cloglog", 
     col = c("blue", "red", "darkgreen"), 
     lty = 1:3,
     lwd = 2,
     main = "Log-Log Survival Plot for PH Assumption",
     xlab = "Log(Time) in Weeks", 
     ylab = "log(-log(Survival))")

# 3. Add legend
legend("topleft", 
       legend = c("Treatment 1", "Treatment 2", "Treatment 3"), 
       col = c("blue", "red", "darkgreen"), 
       lty = 1:3, 
       lwd = 2)
```

Its hard to say if its all that parallel because the steps are so large but I would say it is roughly decently parallel. The proportional hazard model assumption likely can hold here. 

\newpage

## Problem 4

> Consider the survival model characterized by the following piecewise-constant hazard function. where $\pi_1$ and $\pi_2$ are known constants. Based on right-censored data $(x_i,\delta_i)$, i = 1,...,n derive the MLEs of $\lambda_1$,  $\lambda_2$, and  $\lambda_3$ and their standard errors.

This is our assumption:
$$\Lambda(t) = \begin{cases} \lambda_1 t & 0 \le t < \pi_1 \\ \lambda_1 \pi_1 + \lambda_2(t - \pi_1) & \pi_1 \le t < \pi_2 \\ \lambda_1 \pi_1 + \lambda_2(\pi_2 - \pi_1) + \lambda_3(t - \pi_2) & t \ge \pi_2 \end{cases}$$

We aren't given scaling or anything like that so we should assume we are dealing with a model more similar to exponential distribution. 

Then the likelihood for one observation $i$ is $L_i = [\lambda(x_i)]^{\delta_i} S(x_i)$. We expand:
$$L_i = [\lambda(t_i)]^{\delta_i} \exp\left( -\int_0^{x_i} \lambda(u) du \right)$$
The log likelihood is simply:
$$\ln L_i = \delta_i \ln \lambda(t_i) - \int_0^{t_i} \lambda(u) du$$

We can further simplify because this is a piecewise constant hazard so we simplify $S(t) = \exp(-\int_0^t \lambda(u) du)$ with the idea that the cumulative hazard $\Lambda(t) = \int_0^t \lambda(u) du$ is the sum of the hazards across the intervals the individual has survived. Ie its piecewise so let $t_{ij}$ be the time spent by individual $i$ in interval $j$, and $d_{ij}$ be an indicator if individual $i$ died in interval $j$. Then 
$$\int_0^{x_i} \lambda(u) du = \lambda_1 t_{i1} + \lambda_2 t_{i2} + \lambda_3 t_{i3}$$


Now we do this over all the individuals to find the log-likelihood:
$$\ell(\lambda_1, \lambda_2, \lambda_3) = \sum_{i=1}^n \sum_{j=1}^3 (d_{ij} \ln \lambda_j - \lambda_j t_{ij})$$

Let total time at risk in interval $j$: $T_j = \sum_{i=1}^n t_{ij}$ and total events in interval $j$: $D_j = \sum_{i=1}^n d_{ij}$. Importnat to note here that we are counting for time spent inside the interval. If we only know total time alive, we can easily calculate for interval 1: $t_{i1} = \min(t_i, \pi_1)$, for interval 2: $t_{i2} = \max(0, \min(t_i, \pi_2) - \pi_1)$, for interval 3: $t_{i3} = \max(0, t_i - \pi_2)$. 

We simplify further, swap the summation order:
$$\ell(\lambda_1, \lambda_2, \lambda_3) = \sum_{j=1}^3 \left[ \left( \sum_{i=1}^n d_{ij} \right) \ln \lambda_j - \lambda_j \left( \sum_{i=1}^n x_{ij} \right) \right] = \sum_{j=1}^3 (D_j \ln \lambda_j - \lambda_j T_j)$$

Ok, we have simplified formula. 
Now, to find the MLE for each $\lambda_j$, we take the partial derivative with respect to $\lambda_j$ and we set it to zero:
$$\frac{\partial \ell}{\partial \lambda_j} = \frac{D_j}{\lambda_j} - T_j = 0 \implies \hat{\lambda}_j = \frac{D_j}{T_j}$$

This is standard MLE of exponential distribution. 


Finally, the standard error. The terms are sums of independent terms of each $\lambda_i$ so they should be independent. We can simply solve for the variance based on Fisher information number:

Second derivative (based on previous calculations):
$$\frac{\partial^2 \ell}{\partial \lambda_j^2} = -\frac{D_j}{\lambda_j^2}$$

Negative inverse:
$$\widehat{Var}(\hat{\lambda}_j) = \left( -\frac{\partial^2 \ell}{\partial \lambda_j^2} \bigg|_{\hat{\lambda}_j} \right)^{-1} = \frac{\hat{\lambda}_j^2}{D_j} = \frac{(D_j/T_j)^2}{D_j} = \frac{D_j}{T_j^2}$$

Finally, the standard error is the square root:
$$SE(\hat{\lambda}_j) = \frac{\sqrt{D_j}}{T_j}$$


